Secure Composition of Robust and Optimising Compilers

### 2023-06-22 popl/main.tex
Put langs into their own section. Same with comps

@@

\section{Case Study: Memory Safe and Cryptographic Constant Time Code}\label{sec:casestud}
\begin{figure}[h]
  \begin{center}
    \begin{tikzpicture}
      \node (S) {$\src{L_{\tmssafe}}$};
      \node[right=1.0 of S] (T) {$\trg{L}$};
      \node[right=1.0 of T] (M) {$\irl{L_{\mssafe}}$};
      \node[below right=0.5 and 1.0 of M] (D0) {$\irl{L_{\mssafe}}$};
      \node[above right=0.5 and 1.0 of M] (C0) {$\irl{L_{\mssafe}}$};
      \node[right=1.0 of D0] (D1) {$\irl{L_{\mssafe}}$};
      \node[right=1.0 of C0] (C1) {$\irl{L_{\mssafe}}$};
      \node[right=4.5 of M] (O) {$\obj{L_{\scctsafe}}$};

      \draw[dashed] ($(sectms)-(0.7,0)$) -- ($(sectms)-(0.7,1.75)$);
      \draw[dashed] ($(sectms)-(-0.7,0)$) -- ($(sectms)-(-0.7,1.75)$);
      \draw[dashed] ($(secsms)-(-0.7,0)$) -- ($(secsms)-(-0.7,1.75)$);
      \draw[dashed] ($(secscct)-(0.7,0)$) -- ($(secscct)-(0.7,0.75)$);
      \draw[dashed] ($(secscct)-(-0.7,0)$) -- ($(secscct)-(-0.7,0.75)$);
    \end{tikzpicture}
  \end{center}
  \caption{Visualisation of the optimizing compilation pipeline that attains a combination of \gls{ms} and \gls{cct}.}\label{fig:pipeline}
\end{figure}

This section defines several secure compilers, each of which robustly preserves a different property of interest (\Cref{fig:pipeline}).
The power of the framework (\Cref{sec:compprop,sec:compcomp}) is demonstrated by composing these compilers for a secure, but optimizing, compilation chain that robustly preserves \gls{ms} and \gls{cct}.
Concretely, the section starts with a base-line definition for the used languages (\Cref{subsec:cs:defs}) and extends that to define languages $\src{L_{TMS}}$ and $\trg{L}$.
For programmers using $\src{L_{TMS}}$, their mental model should be that their code is temporal memory safe and cryptographic constant time.
The former is proven to be enforced by the static semantics of $\src{L_{TMS}}$, where each well-typed program is temporal memory safe.
The latter is completely untyped.
Having languages defined, this section continues with a compiler $\cc{\src{L_{\tmssafe}}}{\trg{L}}$ and proves that $\cc{\src{L_{\tmssafe}}}{\trg{L}}$ robustly preserves \gls{tms} (\Cref{subsec:cs:tms}).
For readability purposes, it defines $\irl{L_{MS}}$ to be exactly the same as $\trg{L}$ and continues to present a compiler $\cc{\trg{L}}{\irl{L_{MS}}}$ that robustly preserves \gls{sms} (\Cref{subsec:cs:ms}).
Now leaveraging the power of the framework that this paper introduces (\Cref{sec:compcomp,sec:compprop}), a compiler $\rtp{\cc{\src{L_{TMS}}}{\irl{L_{MS}}}}{\mssafe}$ is attained.
The next steps in the compilation pipeline are two secure compilers $\cc[_{\gls{dce}}]{\irl{L_{MS}}}{\irl{L_{MS}}}$ and $\cc[_{\gls{cf}}]{\irl{L_{MS}}}{\irl{L_{MS}}}$ that perform \glsfirst{dce} and \glsfirst{cf}, respectively, to model an optimzing pipeline happening somewhere in the compilation chain (\Cref{subsec:cs:opts}).
Finally, this section defines the language $\obj{L_{\scctsafe}}$ that is, in contrast to the other languages, not cryptographic constant time by definition.
Concluding, a compiler $\cc{\irl{L_{MS}}}{\obj{L_{CCT}}}$ that robustly preserves \gls{cct} (\Cref{subsec:cs:scct}) is shown and composed with $\cc{\src{L_{TMS}}}{\irl{L_{MS}}}$ and the optimizing compilers $\cc[_{\gls{dce}}]{\irl{L_{MS}}}{\irl{L_{MS}}}$ and $\cc[_{\gls{cf}}]{\irl{L_{MS}}}{\irl{L_{MS}}}$.
In turn, this yields $\cc{\src{L_{TMS}}}{\obj{L_{CCT}}}$, a compiler that robustly preserves a combination of \gls{ms} and \gls{cct}.

\subsection{Preliminary Definitions}\label{subsec:cs:defs}
\begin{gather*}
  \begin{aligned}
  &\mi{(Finals)}~\finalexpr\bnfdef\ \valueexpr \mid x\hspace{1.5cm}
  \mi{(Values)}~\valueexpr\bnfdef\ n\\
  \mi{(Expressions)}~\expr&\bnfdef\ \finalexpr \mid \lbinop{\expr[_{1}]}{\expr[_{2}]} \mid \lget{x}{\expr} \mid \lnew{x}{\expr[_{1}]}{\expr[_{2}]} \mid \ldelete{x} \mid \lset{x}{\expr[_{1}]}{\expr[_{2}]} \\
    & \mid \llet{x}{\expr[_{1}]}{\expr[_{2}]} \mid \lreturn{\expr} \mid \lcall{g}{\expr} \mid \lifz{\expr}{\expr[_{1}]}{\expr[_{2}]} \mid \labort \\
  \mi{(State)}~\statevar&\bnfdef \tup{\cfstate}{\sandboxtag,\memstate} \hspace{0.5cm}
  \mi{(Runtime~Terms)}~\runtimetermvar\bnfdef\rtt{\statevar}{\expr} \mid \lightning \hspace{0.5cm}
    \text{where}\ \lbinop{}{}\in\left\{+,-,\times,<,/\right\}\\
  \mi{(Symbols)}~\function&\bnfdef\ \lfunction{g}{x}{e}\hspace{0.5cm}
  \mi{(Types)}~\type\bnfdef\ \natt\hspace{0.5cm}
  \mi{(Locations)}~\loc\in\nat \\
  \mi{(Eval.~Ctx.)}~\ectx&\bnfdef\ \hole{\cdot} \mid \lbinop{\ectx}{\expr} \mid \lbinop{\valueexpr}{\ectx} \mid \lget{x}{\ectx} \mid \lnew{x}{\ectx}{\expr} \mid \lset{x}{\ectx}{\expr}\\
    & \mid \lset{x}{\valueexpr}{\ectx} \mid \lreturn{\ectx} \mid \lcall{g}{\ectx} \mid \lifz{\ectx}{\expr[_{1}]}{\expr[_{2}]} \\
  \mi{(Poison)}~\poison&\bnfdef\ \poisoned \mid \poisonless \hspace{0.5cm}
  \mi{(Store)}~\store\bnfdef\ \hole{\cdot} \mid \storeel{x}{\loc}{\sandboxtag}{\poison}{n},\store \\
  \mi{(Transfer~Tag)}~\comm&\bnfdef\ \nocomm \mid\ \ctxtocomp \mid\ \comptoctx \hspace{0.5cm}
  \mi{(Heaps)}~\heap\bnfdef\ \hole{\cdot} \mid n,\heap \hspace{0.5cm}
  \mi{(Library)}~\library\bnfdef\ \hole{\cdot} \mid \function,\library \\
  \mi{(Component~Names)}~\commlib&\bnfdef\ \hole{\cdot} \mid g,\commlib \hspace{0.5cm}
  \mi{(Continuation~Stack)}~\ectxstack\bnfdef\ \hole{\cdot} \mid \tup{\ectx}{g},\ectxstack \\
  \mi{(Flow~State)}~\cfstate&\bnfdef\ \tup{\commlib}{\library,\ectxstack} \hspace{0.5cm}
  \mi{(Memory~State)}~\memstate\bnfdef\ \tup{\heap[^{\ctx}]}{\heap[^{\comp}],\store} \\
  \mi{(Programs)}~&\prog{\library_{\ctx}}{\library_{\comp}}\hspace{0.5cm}
  \mi{(Substitutions)}~\substvar\bnfdef \hole{\cdot} \mid \subst{\valueexpr}{x},\gamma \\
  \mi{(Pre\text{--}Events)}~\event[_{b}]&\bnfdef \ev{Alloc}\ \loc\ \valueexpr \mid \ev{Dealloc}\ \loc \mid \ev{Get}\ \loc\ \valueexpr \mid \ev{Set}\ \loc\ \valueexpr\ \valueexpr['] \mid \ev{Call}\ \comm\ g\ \valueexpr \mid \ev{Ret}\ \comm\ \valueexpr\\
    & \mid \ev{Start} \mid \ev{End}\ \valueexpr\hspace{0.5cm}
  \mi{(Events)}~\event\bnfdef \emptyevent \mid \lightning \mid \tup{\event[_{b}]}{\sandboxtag} \\
  \end{aligned}
\end{gather*}

Above is the base syntax of all the programming languages of this paper.
Expressions ($\expr$) can only be evaluated in the form of runtime--terms ($\rtt{\statevar}{\expr}$).
States ($\statevar$) have a flow- ($\cfstate$) and a memory state ($\memstate$).
As for the flow state, the list of component--level names ($\commlib$), the list of symbols ($\library$), and a stack of continuations tagged with the name of a symbol ($\ectxstack$) are necessary for calling and returning while emitting the right transfer tag ($\sandboxtag$) depending on whether there is a context switch from context to component ($\comptoctx$), from component to context ($\ctxtocomp$), or whether control flow stays internal to either ($\nocomm$).
The memory state contains both a context--level ($\heap[^{\ctx}]$) and a component--level heap ($\heap[^{comp}]$) as well as a store ($\store$).
Effectively, it is syntactically impossible for a context--level runtime--term to have a pointer into the component--level heap and vice--versa.
Pointers are non-nullable and there is no syntactic way to let them point into the middle of an object.
However, pointers always point to arrays and the individual elements of that array are read- and writeable by means of adding an offset that may be negative.
Heap locations are second--class. %The syntax is constrained as such that one cannot simply guess a heap location.
Stores map identifiers to pointer metadata.
This metadata is a tuple with the location ($\loc$), the control--tag ($\sandboxtag$), poison--tag ($\poison$), and the size of the allocated object ($n$).
The control-tag allows to distinguish whether a location indexes into $\heap[^{\ctx}]$ or $\heap[^{\comp}]$.
The poison--tag marks a location as freed ($\poisoned$) without removing the entry from $\store$.
If one removes an entry, for example, by evaluating $\ldelete{x}$, then there is no way to distinguish between freed pointers and free variables.
The latter should render evaluation as stuck, while the former should emit a $\ev{Dealloc}\ \loc$ event regardless of wether the object a pointer points to is still allocated or not.
Programs ($\prog{\library_{\ctx}}{\library_{\comp}}$) have disjoint sets of symbols, one for context-level ($\library_{\ctx}$) and one for component-level symbols ($\library_{\comp}$).

All programming languages presented in this section have an environmental semantics and all but $\src{L_{\tmssafe}}$ are dynamically typed.
Each evaluation may emit the empty event ($\emptyevent$), a program crash token ($\lightning$), or a pair containing a pre--event ($\event[_{b}]$) and the corresponding control--tag ($\sandboxtag$) indicating whether the event fired during an evaluation of a context or a component.
Note that the reflexive-transitive closure used in this paper behaves like a big-step semantics, where the reflexive case is ony for the empty trace and finals, and the emitted trace does not contain empty events.
Most rules are entirely standard and omitted.
Heaps may need to grow when allocating, which the section writes as $\growh{\heap}{n}$, denoting that $\heap$ grows by $n$ cells.
Indexing into a heap, looking up symbols in libraries, or accessing the metadata of a pointer from a store are all written as a function call.
For example, the value located at the fifth position in heap $\heap$ is written as $\heap(5)$.
The operation to set the fifth position in heap $\heap$ to value $\valueexpr$ is written as $\seth{\heap}{5}{\valueexpr}$ and yields a new heap that is entirely the same to $\heap$ modulo the fifth position.

@@

\subsection{$\src{L_{\tmssafe}}$: A Temporal but Not Spatial Memory Safe Language}\label{subsec:ltms}
\begin{gather*}
  \begin{aligned}
    \mi{(Pre\text{--}Types)}~\src{\type[_{e}]}\bnfdef&\ \src{\natt} \mid \src{\ptrn} &
    \mi{(Type~Environment)}~\src{\typenv}\bnfdef&\ \src{\hole{\cdot}} \mid \hastype{\src{x}}{\src{\type}},\src{\typenv} \\
    \mi{(Types)}~\src{\type}\bnfdef&\ \src{\type[_{e}]} \mid \src{\type[_{e}]\to\bot} \mid \src{\type[^{(1)}_{e}]\to\type[^{(2)}_{e}]} &
    \mi{(Pointer\ Qualifiers)}~\src{\ptrqual}\bnfdef&\ \src{\fullq} \mid \src{\halfq} \\
    \mi{(Symbols)}~\src{\function}\bnfdef&\src{\lfunction{g}{x:\type[_{e}]\to\type[_{e}]}{\expr}} \\
  \end{aligned}
\end{gather*}
\begin{center}
  \judgbox{\tcheck{\src{\typenv}}{\src{\expr}}{\src{\type}}}{,,Under environment $\src{\typenv}$, expression $\src{\expr}$ has type $\src{\type}$.''}
  %
  \typerule{t-$\src{x}$}{
    \src{\typenv}(\src{x})=\src{\type} &
    \notowned{\src{\typenv}}
  }{
    \tcheck{\src{\typenv}}{\src{x}}{\src{\type}}
  }{t-x}
  %
  \typerule{t-$\src{get}$}{
    \tcheck{\src{\typenv[_2]}}{\src{x}}{\src{\wptr}} &
    \tcheck{\src{\typenv[_1]}}{\src{\expr}}{\src{\natt}} &
  }{
    \tcheck{\typenvsplit{\src{\typenv[_1]}}{\src{\typenv[_2]}}}{\src{\lget{x}{\expr}}}{\src{\natt}}
  }{t-get}
  %
  \typerule{t-$\src{let}$}{
    \tcheck{\src{\typenv[_1]}}{\src{\expr[_1]}}{\src{\type[_e]}} &
    \tcheck{\hastype{\src{x}}{\src{\type[_e]}},\src{\typenv[_2]}}{\src{\expr[_2]}}{\src{\type}}
  }{
    \tcheck{\typenvsplit{\src{\typenv[_1]}}{\src{\typenv[_2]}}}{\src{\llet{x}{\expr[_1]}{\expr[_2]}}}{\src{\type}}
  }{t-let}
  %
  \typerule{t-$\src{delete}$}{
    \src{\typenv}(\src{x})=\src{\ptr} &
    \notowned{\src{\typenv}\setminus\left\{\hastype{\src{x}}{\src{\ptr}}\right\}}
  }{
    \tcheck{\src{\typenv}}{\src{\ldelete{x}}}{\src{\natt}}
  }{t-delete}
  %
  \typerule{t-$\src{abort}$}{
    \notowned{\src{\typenv}}
  }{
    \tcheck{\src{\typenv}}{\src{\labort}}{\src{\type}}
  }{t-abort}
  %
  \typerule{t-$\src{call}$}{
    \inttype{\src{\type[^{(1)}_e]}} &
    \inttype{\src{\type[^{(2)}_e]}} \\
    \tcheck{\src{\typenv}}{\src{foo}}{\src{\type[^{(1)}_e]\to\type[^{(2)}_e]}} &
    \tcheck{\src{\typenv}}{\src{\expr}}{\src{\type[^{(1)}_e]}}
  }{
    \tcheck{\src{\typenv}}{\src{\lcall{foo}{\expr}}}{\src{\type[^{(2)}_e]}}
  }{t-call}
  %
  \typerule{t-$\src{return}$}{
    \inttype{\src{\type[_e]}} &
    \tcheck{\src{\typenv}}{\src{\expr}}{\src{\type[_e]}}
  }{
    \tcheck{\src{\typenv}}{\src{\lreturn{\expr}}}{\src{\type[_e]\to\bot}}
  }{t-return}
\end{center}

@@

\MKin{show at least one forward sim lemma or not?}

@@

\subsection{$\trg{L}$: An Unsafe Language}\label{subsec:lsms}

@@

The $\trg{L}$ language does not provide any guarantees with regards to memory safety.
The syntax of it extends the unified syntax for all languages (\Cref{subsec:cs:defs}) with pairs of natural numbers (\Cref{tr:e-proj1}) and two different kinds of dynamic checks.
One of those checks allows programmers to investigate wether a pointer is freed or not (\Cref{tr:e-ispoisoned-yes}).
The other check is a simple dynamic type check that evaluates to $\trg{0}$ if the check suceeds and to $\trg{1}$ otherwise (\Cref{tr:e-n-has-nat,tr:e-n-has-nattimesnat}).
The projection $\filterL{\trace}$ is entirely similar to $\filterLtms{\trace}$.

>>>>

\DeclareMathAlphabet{\mathtt}{OT1}{lmtt}{m}{n}
\SetMathAlphabet{\mathtt}{bold}{OT1}{lmtt}{bx}{n}

@@

%\section{Case Study: Memory Safe and Cryptographic Constant Time Code}\label{sec:casestud}
\section{Case Study: Preliminary Definitions}\label{sec:casestud:defs}

In order to define secure compilers one needs programming languages.
To this end, this section defines the languages $\Ltms$, $\Ltrg$, $\Lms$, and $\Lscct$.
In this paper, $\Ltms$ is the only statically typed language and exhibits the property that all well-typed programs are \gls{tms}.
However, one can still write and evaluate $\Ltms$ programs that are well-typed, but perform out-of-bounds memory accesses, so the language is not \gls{sms}.
In contrast, $\Ltrg$ is untyped and does not provide any guarantees with regards to \gls{ms}.
$\Lms$ is exactly the same language as $\Ltrg$, but this paper still distinguishes the two for sake of readability.
For all these languages, so $\Ltms$, $\Ltrg$, and $\Lms$, programmers assume \gls{cct} to hold, because the compiler is said to ensure it, similarily to what is being done in literature~\cite{cauligi2019fact}.
This assumption does not hold anymore for the language $\Lscct$, which leaks information on branches and binary operations.
But, contrary to the other languages, one can read and write to a model-specific register indicating a ``\gls{cct}-mode''.

\subsection{Shared Language Definitions}\label{subsec:cs:defs}
\begin{gather*}
  \begin{aligned}
  &\mi{(Finals)}~\finalexpr\bnfdef\ \valueexpr \mid x \hspace{0.5cm}
  \mi{(Values)}~\valueexpr\bnfdef\ n \hspace{0.5cm}
  \mi{(Types)}~\type\bnfdef\ \natt \\
  \mi{(Expressions)}~\expr&\bnfdef\ \finalexpr \mid \lbinop{\expr[_{1}]}{\expr[_{2}]} \mid \lget{x}{\expr} \mid \lnew{x}{\expr[_{1}]}{\expr[_{2}]} \mid \ldelete{x} \mid \lset{x}{\expr[_{1}]}{\expr[_{2}]} \\
    & \mid \llet{x}{\expr[_{1}]}{\expr[_{2}]}  \mid \lreturn{\expr} \mid \lcall{g}{\expr} \mid \lifz{\expr}{\expr[_{1}]}{\expr[_{2}]} \mid \labort \\
  \mi{(Symbols)}~\function&\bnfdef\ \lfunction{g}{x}{e}\hspace{0.5cm}
  \mi{(Libraries)}~\library\bnfdef\ \hole{\cdot} \mid \function,\library \hspace{1.00cm}
    \text{where}\ \lbinop{}{}\in\left\{+,-,\times,<,/\right\}\\
  &\mi{(Programs)}~\prog{\library_{\ctx}}{\library_{\comp}}\hspace{0.5cm}
  \mi{(Substitutions)}~\substvar\bnfdef \hole{\cdot} \mid \subst{\valueexpr}{x},\gamma \\
  \end{aligned}
\end{gather*}

Above is the base syntax of all the programming languages of this paper.
\MK{this needs a complete rewrite}
Expressions ($\expr$) can only be evaluated in the form of runtime--terms ($\rtt{\statevar}{\expr}$).
States ($\statevar$) have a flow- ($\cfstate$) and a memory state ($\memstate$).
As for the flow state, the list of component--level names ($\commlib$), the list of symbols ($\library$), and a stack of continuations tagged with the name of a symbol ($\ectxstack$) are necessary for calling and returning while emitting the right transfer tag ($\sandboxtag$) depending on whether there is a context switch from context to component ($\comptoctx$), from component to context ($\ctxtocomp$), or whether control flow stays internal to either ($\nocomm$).
The memory state contains both a context--level ($\heap[^{\ctx}]$) and a component--level heap ($\heap[^{comp}]$) as well as a store ($\store$).
Effectively, it is syntactically impossible for a context--level runtime--term to have a pointer into the component--level heap and vice--versa.
Pointers are non-nullable and there is no syntactic way to let them point into the middle of an object.
However, pointers always point to arrays and the individual elements of that array are read- and writeable by means of adding an offset that may be negative.
Heap locations are second--class. %The syntax is constrained as such that one cannot simply guess a heap location.
Stores map identifiers to pointer metadata.
This metadata is a tuple with the location ($\loc$), the control--tag ($\sandboxtag$), poison--tag ($\poison$), and the size of the allocated object ($n$).
The control-tag allows to distinguish whether a location indexes into $\heap[^{\ctx}]$ or $\heap[^{\comp}]$.
The poison--tag marks a location as freed ($\poisoned$) without removing the entry from $\store$.
If one removes an entry, for example, by evaluating $\ldelete{x}$, then there is no way to distinguish between freed pointers and free variables.
The latter should render evaluation as stuck, while the former should emit a $\ev{Dealloc}\ \loc$ event regardless of wether the object a pointer points to is still allocated or not.
Programs ($\prog{\library_{\ctx}}{\library_{\comp}}$) have disjoint sets of symbols, one for context-level ($\library_{\ctx}$) and one for component-level symbols ($\library_{\comp}$).

\begin{gather*}
  \begin{aligned}
  \mi{(Eval.~Ctx.)}~\ectx&\bnfdef\ \hole{\cdot} \mid \lbinop{\ectx}{\expr} \mid \lbinop{\valueexpr}{\ectx} \mid \lget{x}{\ectx} \mid \lnew{x}{\ectx}{\expr} \mid \lset{x}{\ectx}{\expr}\\
    & \mid \lset{x}{\valueexpr}{\ectx} \mid \lreturn{\ectx} \mid \lcall{g}{\ectx} \mid \lifz{\ectx}{\expr[_{1}]}{\expr[_{2}]} \\
  \end{aligned}\\
  \begin{aligned}
  \mi{(States)}~\statevar&\bnfdef \tup{\cfstate}{\sandboxtag,\memstate} &
  \mi{(Poison)}~\poison&\bnfdef\ \poisoned \mid \poisonless &
  \mi{(Locations)}~\loc\in\nat \\
  \end{aligned}\\
  \begin{aligned}
  \mi{(Heaps)}~\heap&\bnfdef\ \hole{\cdot} \mid n,\heap &
  \mi{(Stores)}~\store&\bnfdef\ \hole{\cdot} \mid \storeel{x}{\loc}{\sandboxtag}{\poison}{n},\store &
  \mi{(Transfer~Tags)}~\comm\bnfdef\ \nocomm \mid\ \ctxtocomp \mid\ \comptoctx \\
  \end{aligned}\\
  \begin{aligned}
  \mi{(Runtime~Terms)}~\runtimetermvar&\bnfdef\rtt{\statevar}{\expr} \mid \lightning &
  \mi{(Component~Names)}~\commlib&\bnfdef\ \hole{\cdot} \mid g,\commlib \\
  \end{aligned}\\
  \begin{aligned}
  \mi{(Continuation~Stacks)}~&\ectxstack\bnfdef\ \hole{\cdot} \mid \tup{\ectx}{g},\ectxstack \\
  \end{aligned}
\end{gather*}
All programming languages presented in this section have an environmental semantics and all but $\src{L_{\tmssafe}}$ are dynamically typed.
Each evaluation may emit the empty event ($\emptyevent$), a program crash token ($\lightning$), or a pair containing a pre--event ($\event[_{b}]$) and the corresponding control--tag ($\sandboxtag$) indicating whether the event fired during an evaluation of a context or a component.
Note that the reflexive-transitive closure used in this paper behaves like a big-step semantics, where the reflexive case is ony for the empty trace and finals, and the emitted trace does not contain empty events.
Most rules are entirely standard and omitted.
Heaps may need to grow when allocating, which the section writes as $\growh{\heap}{n}$, denoting that $\heap$ grows by $n$ cells.
Indexing into a heap, looking up symbols in libraries, or accessing the metadata of a pointer from a store are all written as a function call.
For example, the value located at the fifth position in heap $\heap$ is written as $\heap(5)$.
The operation to set the fifth position in heap $\heap$ to value $\valueexpr$ is written as $\seth{\heap}{5}{\valueexpr}$ and yields a new heap that is entirely the same to $\heap$ modulo the fifth position.

\begin{gather*}
  \begin{aligned}
  \mi{(Pre\text{--}Events)}~\event[_{b}]\bnfdef \ev{Alloc}\ \loc\ \valueexpr \mid \ev{Dealloc}\ \loc \mid \ev{Get}\ \loc\ \valueexpr &\mid \ev{Set}\ \loc\ \valueexpr\ \valueexpr['] \mid \ev{Call}\ \comm\ g\ \valueexpr \mid \ev{Ret}\ \comm\ \valueexpr \mid \ev{Start} \mid \ev{End}\ \valueexpr \\
  \mi{(Events)}~\event&\bnfdef \emptyevent \mid \lightning \mid \tup{\event[_{b}]}{\sandboxtag} \\
  \end{aligned}
\end{gather*}

\MKin{now is time to talk about events? or maybe right after presenting the rules}

@@

\subsection{$\src{L_{\tmssafe}}$: A Temporal but Not Spatial Memory Safe Language}\label{subsec:ltms}
\begin{gather*}
  \begin{aligned}
    \mi{(Pre\text{--}Types)}~\src{\type[_{e}]}\bnfdef&\ \src{\natt} \mid \src{\ptrn} &
    \mi{(Type~Environments)}~\src{\typenv}\bnfdef&\ \src{\hole{\cdot}} \mid \hastype{\src{x}}{\src{\type}},\src{\typenv} \\
    \mi{(Types)}~\src{\type}\bnfdef&\ \src{\type[_{e}]} \mid \src{\type[_{e}]\to\bot} \mid \src{\type[^{(1)}_{e}]\to\type[^{(2)}_{e}]} &
    \mi{(Pointer\ Qualifiers)}~\src{\ptrqual}\bnfdef&\ \src{\fullq} \mid \src{\halfq} \\
    \mi{(Symbols)}~\src{\function}\bnfdef&\src{\lfunction{g}{x:\type[_{e}]\to\type[_{e}]}{\expr}} \\
  \end{aligned}
\end{gather*}
\begin{center}
  \judgbox{\tcheck{\src{\typenv}}{\src{\expr}}{\src{\type}}}{,,Under environment $\src{\typenv}$, expression $\src{\expr}$ has type $\src{\type}$.''}
  %
  \typerule{t-$\src{x}$}{
    \src{\typenv}(\src{x})=\src{\type} &
    \notowned{\src{\typenv}}
  }{
    \tcheck{\src{\typenv}}{\src{x}}{\src{\type}}
  }{t-x}
  %
  \typerule{t-$\src{get}$}{
    \tcheck{\src{\typenv[_2]}}{\src{x}}{\src{\wptr}} &
    \tcheck{\src{\typenv[_1]}}{\src{\expr}}{\src{\natt}} &
  }{
    \tcheck{\typenvsplit{\src{\typenv[_1]}}{\src{\typenv[_2]}}}{\src{\lget{x}{\expr}}}{\src{\natt}}
  }{t-get}
  %
  \typerule{t-$\src{let}$}{
    \tcheck{\src{\typenv[_1]}}{\src{\expr[_1]}}{\src{\type[_e]}} &
    \tcheck{\hastype{\src{x}}{\src{\type[_e]}},\src{\typenv[_2]}}{\src{\expr[_2]}}{\src{\type}}
  }{
    \tcheck{\typenvsplit{\src{\typenv[_1]}}{\src{\typenv[_2]}}}{\src{\llet{x}{\expr[_1]}{\expr[_2]}}}{\src{\type}}
  }{t-let}
  %
  \typerule{t-$\src{delete}$}{
    \src{\typenv}(\src{x})=\src{\ptr} &
    \notowned{\src{\typenv}\setminus\left\{\hastype{\src{x}}{\src{\ptr}}\right\}}
  }{
    \tcheck{\src{\typenv}}{\src{\ldelete{x}}}{\src{\natt}}
  }{t-delete}
  %
  \typerule{t-$\src{abort}$}{
    \notowned{\src{\typenv}}
  }{
    \tcheck{\src{\typenv}}{\src{\labort}}{\src{\type}}
  }{t-abort}
  %
  \typerule{t-$\src{call}$}{
    \inttype{\src{\type[^{(1)}_e]}} &
    \inttype{\src{\type[^{(2)}_e]}} \\
    \tcheck{\src{\typenv}}{\src{foo}}{\src{\type[^{(1)}_e]\to\type[^{(2)}_e]}} &
    \tcheck{\src{\typenv}}{\src{\expr}}{\src{\type[^{(1)}_e]}}
  }{
    \tcheck{\src{\typenv}}{\src{\lcall{foo}{\expr}}}{\src{\type[^{(2)}_e]}}
  }{t-call}
  %
  \typerule{t-$\src{return}$}{
    \inttype{\src{\type[_e]}} &
    \tcheck{\src{\typenv}}{\src{\expr}}{\src{\type[_e]}}
  }{
    \tcheck{\src{\typenv}}{\src{\lreturn{\expr}}}{\src{\type[_e]\to\bot}}
  }{t-return}
\end{center}

@@

@@

\subsection{$\Ltrg$: A Memory-Unsafe Language}\label{subsec:lsms}

@@

The $\Ltrg$ language does not provide any guarantees with regards to memory safety.
The syntax of it extends the unified syntax for all languages (\Cref{subsec:cs:defs}) with pairs of natural numbers (\Cref{tr:e-proj1}) and two different kinds of dynamic checks.
One of those checks allows programmers to investigate wether a pointer is freed or not (\Cref{tr:e-ispoisoned-yes}).
The other check is a simple dynamic type check that evaluates to $\trg{0}$ if the check suceeds and to $\trg{1}$ otherwise (\Cref{tr:e-n-has-nat,tr:e-n-has-nattimesnat}).
The projection $\filterL{\trace}$ is entirely similar to $\filterLtms{\trace}$.
\MKin{add note about \gls{sms} projection}

\subsection{$\Lms$: Another Memory-Unsafe Language}\label{subsec:lms}

This paper chooses to introduce $\Lms$, despite it being exactly equal to $\Ltrg$.
The main purpose of this decision is to enhance readability.

\MKin{add note about \gls{sms} projection}

\subsection{$\Lscct$: An Unsafe Language}\label{subsec:lscct}

\begin{gather*}
  \begin{aligned}
    \mi{(Expressions)}~\obj{\expr}\bnfdef&\dots \mid \obj{\lwrdoit{\expr}} \mid \obj{\lrddoit{x}{\expr}} &
    \mi{(States)}~\obj{\statevar}\bnfdef&\obj{\tup{\cfstate}{\sandboxtag,n,\memstate}}
  \end{aligned}
\end{gather*}

$\Lscct$ extends the state presented earlier (\Cref{subsec:cs:defs}) with another entry that models the value of a register for a data (operand) independent timing mode.
To modify the register, $\Lscct$ extends expressions with a way to write a value to the register ($\obj{\lwrdoit{\expr}}$) and a way to read it ($\obj{\lrddoit{x}{\expr}}$).
Finals carry a secrecy annotation, indicating high ($\obj{\lock}$) or low ($\obj{\unlock}$) secrecy.
Whenever the register is marked as active, parts of the dynamics of $\Lscct$ do not leak information.
In general, $\Lscct$ can leak more information than the other languages:

\begin{gather*}
  \begin{aligned}
    \mi{(Pre\text{--}Events)}~\obj{\event[_{b}]}&\bnfdef\ \dots \mid \obj{{Branch}\ n} \mid \obj{{Binop}\ n} \mid \obj{\widehat{{Get}}\ \loc\ \valueexpr} \mid \obj{\widehat{{Set}}\ \loc\ \valueexpr\ \valueexpr[']} &
    \mi{(Events)}~\obj{\event}&\bnfdef\ \obj{\varepsilon} \mid \obj{\lightning} \mid \obj{\tup{\event[_{b}]}{\sandboxtag,\securitytag}}
  \end{aligned}
\end{gather*}

Events carry a secrecy tag instead of the data they carry.
Branches and binary operations leak information about the data they compute on.
If the data (object) independent timing register is set to be active, the read and write to memory events should not leak.
However, to emit $\obj{\emptyevent}$ in this case would mean that memory unsafe operations are hidden as well.
Since this language should not have a mechanism to hide memory unsafety, pre--events are extended to model reading from ($\obj{\widehat{{Get}}\ \loc\ \valueexpr}$) and writing to ($\obj{\widehat{Set}\ \loc\ \valueexpr}$) memory without leaking secret information involved in the access.

\begin{center}
  \typerule{e-$\obj{\lbinop{}{}}$-leak}{
    \obj{n_1}\oplus\obj{n_2}=\obj{n_3} &
    \obj{\securitytag}\sqcap\obj{\securitytag[']}=\obj{\securitytag['']} &
    \obj{\oplus}\notin\left\{\obj{\times}, \obj{/}, \obj{<}\right\}
  }{
    \pstepto{\rtt{\obj{\tup{\cfstate}{\sandboxtag,0,\memstate}}}{\obj{\lbinop{n_1^{\securitytag}}{n_2^{\securitytag[']}}}}}{\rtt{\obj{\tup{\cfstate}{\sandboxtag,0,\memstate}}}{\obj{n_3^{\securitytag['']}}}}{\obj{\tup{Binop\ n_3}{\sandboxtag,\securitytag['']}}}
  }{e-binop-leak}
\end{center}

\MKin{filters to spec events}

\section{Case Study: Secure Compiler Compositions in the Wild}\label{sec:casestud:rtp}

\begin{figure}[h]
  \begin{center}
    \begin{tikzpicture}
      \node (S) {$\src{L_{\tmssafe}}$};
      \node[right=1.0 of S] (T) {$\trg{L}$};
      \node[right=1.0 of T] (M) {$\irl{L_{\mssafe}}$};
      \node[below right=0.5 and 1.0 of M] (D0) {$\irl{L_{\mssafe}}$};
      \node[above right=0.5 and 1.0 of M] (C0) {$\irl{L_{\mssafe}}$};
      \node[right=1.0 of D0] (D1) {$\irl{L_{\mssafe}}$};
      \node[right=1.0 of C0] (C1) {$\irl{L_{\mssafe}}$};
      \node[right=4.5 of M] (O) {$\obj{L_{\scctsafe}}$};
